---
title: "STAT 335, **Unit 3**: Time Series Regression and Exploratory Data Analysis"
subtitle: "**(Part b): Ordinary Least Squares Regression for Time Series**"
author: "Dr. Mary Lynn Reed"
date: ""
theme: "Warsaw"
colortheme: "wolverine"
fontsize: 10pt
output: beamer_presentation
header-includes:
  - \usefonttheme{serif}
  - \newtheorem{prop}{Proposition}
  - \newtheorem{thm}{Theorem}
  - \newtheorem{cor}{Corollary}
  - \newtheorem{exer}{Exercise}
  - \setbeamertemplate{footline}[frame number]
  - \usepackage{setspace}
  - \newcommand{\R}{\mathbb{R}}
  - \newcommand{\var}{\text{var}}
  - \newcommand{\cov}{\text{cov}}
  - \newcommand{\corr}{\text{corr}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(xts)
library(astsa)
```

## Regression for Time Series

We want to look at the situation where a time series, say, $x_t$, for $t = 1, \ldots, n,$ is possibly being influenced by a collection of other time series, say
$$z_{t1}, z_{t2}, \ldots, z_{tq}.$$
These independent time series are called **exogenous variables** (since they are ``external" to the series $x_t$).

>

\only<2->{
So, we're looking at a {\it linear regression model} of: 
$$x_t = \beta_0 + \beta_1 z_{t1} + \beta_2 z_{t2} + \cdots + \beta_q z_{tq} + w_t,$$
where the $\beta_0, \beta_1, \ldots, \beta_q$ are unknown fixed regression coefficients and $w_t$ is a white noise time series with variance $\sigma_w^2$.
}


## Example: Estimating the Linear Trend of a Commodity

In the **astsa** package is a dataset **salmon** that contains the monthly export price of farm bred Norweigian salmon, from September 2003 to June 2017, in US dollars per kilogram.

```{r, out.height="70%", out.width="70%", fig.align="center"}
tsplot(salmon, main="Export Price of Salmon", ylab="USD per kg")
```


## Example: Estimating the Linear Trend of a Commodity

We want to estimate the *linear trend* of this time series by doing a linear regression on this data and finding the best fit line.

I.e., we'll fit the model:
$$x_t = \beta_0 + \beta_1 z_t + w_t,$$
where: 

- $w_t$ is the error term, which we will assume is modeled by white noise (may not be accurate but we will assume for now)

- $z_t$ will be time, measured in months, since the $x_t$ series is given in monthly increments


## Example: Estimating the Linear Trend of a Commodity

Let's look at the **salmon** time series data set

\begin{block}{Exercise 3b.1}
Make sure the {\tt astsa} package is loaded and then do the following commands to look at the {\tt salmon} dataset:
\vskip .3cm
{\tt salmon}
\vskip .3cm
{\tt time(salmon)}
\vskip .3cm
Given what you see, what are the values for $x_1$ and $z_1$?
\end{block}

## Example: Estimating the Linear Trend of a Commodity

Now, we'll perform linear regression using the R command **lm()**.

```{r}
summary(fit <- lm(salmon ~ time(salmon)))
```

## Example: Estimating the Linear Trend of a Commodity

```{r}
fit <- lm(salmon ~ time(salmon))
tsplot(salmon, col=4, ylab="USD per KG", main="Salmon Export Price")
abline(fit,col="red")
```

## Example: Estimating the Linear Trend of a Commodity

What's the equation of the regression line we fit?
```{r}
(fit <- lm(salmon ~ time(salmon)))
```

So, the ordinary least squares "best fit" regression line is:
$$x_t = -503.0895 + 0.2529 z_t$$


## Example: Estimating the Linear Trend of a Commodity

How good was the assumption that the error term is modeled as white noise? Let's look at the residuals...

```{r, out.height="80%"}
fit <- lm(salmon ~ time(salmon))
par(mfrow=c(1,2))
plot(resid(fit))
qqnorm(resid(fit))
```

## Back to the General Situation...Multiple Regression for Time Series

Suppose $x_t$ is a time series and we're fitting the *linear regression model*
$$x_t = \beta_0 + \beta_1 z_{t1} + \beta_2 z_{t2} + \cdots + \beta_q z_{tq} + w_t$$

where $z_{t1}, z_{t2}, \ldots, z_{tq}$ are independent variables and $\{ w_t \}$ is the error term, which we assume is white noise with variance $\sigma_w^2$.

>

\only<2->{
Ordinary least squares (OLS) regression finds the regression coefficients $\beta_0, \beta_1, \ldots, \beta_q$ that minimize the error sum of squares:
$$S = \sum_{t=1}^n w_t^2 = \sum_{t=1}^n (x_t - [\beta_0 + \beta_1 z_{t1} + \beta_2 z_{t2} + \cdots + \beta_q z_{tq}])^2$$
}

## Multiple Regression for Time Series

The minimized error sum of squares is: 
$$\text{SSE} = \sum_{t=1}^n (x_t - \hat{x_t})^2,$$
where $\hat{x_t} = \hat{\beta_0} + \hat{\beta_1} z_{t1} + \hat{\beta_2} z_{t2} + \cdots + \hat{\beta_q} z_{tq}$ and $\hat{\beta_i}$ is the (OLS) regression estimate of $\beta_i$. 

>

An unbiased estimator for the variance parameter $\sigma_w^2$ is: 
$$s_w^2 = \text{MSE} = \frac{\text{SSE}}{n - (q+1)},$$
where MSE denotes the **mean square error**.

## Standard Error of the Regression Coefficients

We denote the standard error for the estimated regression coefficient $\hat{\beta_i}$ as: $\text{se}(\hat{\beta_i})$. If the errors are normal, the statistic
$$t = \frac{\hat{\beta_i} - \beta_i}{\text{se}(\hat{\beta_i})}$$
will follow a $t$-distribution with $n - (q+1)$ degrees of freedom. 

>

**Why do we care?** Because this can be used to do hypothesis testing on whether the regression coefficient $\beta_i$ is 0...which will tell us whether the corresponding independent variable $z_{ti}$ is really needed in the model.

## Hypothesis Testing to Select Best Subset of Independent Variables

Suppose a proposed model specifies that only a subset $r < q$ independent variables, say
$$\{ z_{t1}, z_{t2}, \ldots, z_{tr}\}$$
is influencing the dependent variable $x_t$. 

>

So we want to devise a test to see if this reduced model
$$x_t = \beta_0 + \beta_1 z_{t1} + \beta_2 z_{t2} + \cdots \beta_r z_{tr} + w_t$$ is adequate.

>

So, we want to do a hyptothesis test against the full model, where the null hypothesis is:
$$H_0 : \beta_{r+1} = \cdots = \beta_q = 0.$$

## Hypothesis Testing to Select Best Subset of Independent Variables

The statistic we use for our hyptothesis test is the $F$-statistic:
$$F = \frac{(\text{SSE}_r - \text{SSE})/(q-r)}{\text{SSE}/(n-q-1)} = \frac{\text{MSR}}{\text{MSE}}$$
where $\text{SSE}_r$ is the error sum of squares under the reduced model and the other error statistics are for the full model.


\only<2->{If the null hypthothesis $$H_0 : \beta_{r+1} = \cdots = \beta_q = 0$$ is true, then $\text{SSE}_r \approx \text{SSE}$. Under this hyptothesis, the $F$-statistic has a central $F$-distribution with $q-r$ and $n-q-1$ degrees of freedom. }

\only<3->{
The hull hypothesis is rejected at level $\alpha$ if $F > F_{n-q-1}^{q-r}(\alpha)$, the $(1-\alpha)$ percentile of the $F$-distribution with $q-r$ numerator and $n-q-1$ denominator degrees of freedom.
}

## Special Case

A special case of interest is 
$$H_0 = \beta_1 = \cdots = \beta_q = 0.$$
In this case the reduced model becomes
$$x_t = \beta_0 + w_t.$$
The residual sum of squares under this reduced model is:
$$\text{SSE}_0 = \sum_{t=1}^n (x_t - \bar{x})^2$$
and $\text{SSE}_0$ is often called $\text{SST}$ (*adjusted total sum of squares*).


Recall from our Regression Refresher that $$R^2 = 1 - \frac{\text{SSE}}{\text{SST}}$$ is the **coefficient of determination**. This measures the proportion of variation accounted for by all the variables (in the original model).

## Stepwise Regression

- In statistical modeling, in general, we always seek to find the most *accurate* model with the least amount of *complexity*. This follows the principle of *parsimony* (or *Occam's razor*).

>

- For regression, this means we're looking for the model that has the best fit with the fewest number of parameters. 

>

- We can use the hypothesis testing approach, in a systematic manner, to leave variables out and measure the impact on accuracy and fit. 

## Accuracy and Complexity Measures

- We use the error sum of squares 
$$\text{SSE} = \sum_{t=1}^n (x_t - \hat{x}_t)^2$$
to measure **accuracy** of a model since it measures how close the fitted values ($\hat{x}_t$) are to the actual data ($x_t$).

>

- For a normal regression model with $k$ coefficients, we also have the maximum likelihood estimator for the variance as:
$$\hat{\sigma}_k^2 = \frac{\text{SSE}(k)}{n},$$
where $\text{SSE}(k)$ indicates the residual sum of squares under the model with $k$ regression coefficients. The **complexity** of the model can be characterized by $k$, the number of parameters in the model.

## Akaike's Information Criteria (AIC)

- Seeking to balance the accuracy of the fit of a model against the number of parameters in the model, we look to the notion of an **information criteria**. Here's one (due to Akaike, 1974):

>

\begin{definition}[Akaike's Information Criterion (AIC)]
$$\text{AIC} = \log \hat{\sigma}_k^2 + \frac{n+2k}{n}.$$
\end{definition}

>

- We then note that a *parsimonious model* will be an accurate one (with small error $\hat{\sigma}_k^2$) that is not overly complex (small $k$). So we seek to minimize the AIC to specify the "best" model using this criteria.

>

- Note that "AIC" stands for "An Information Criterion"



## Other Information Criteria 

- Finding the "right" penalty term in the information criteria is a subject of some debate. So here are a few alternative criteria:

>

\begin{definition}[AIC, Bias Corrected(AICc)]
$$\text{AICc} = \log \hat{\sigma}_k^2 + \frac{n+k}{n-k-2}.$$
\end{definition}

>

\begin{definition}[Bayesian Information Criterion, (BIC)]
$$\text{BIC} = \log \hat{\sigma}_k^2 + \frac{k\log n}{n}.$$
\end{definition}

The BIC is also called the **Schwarz Information Criterion (SIC).**

- Studies indicate that the BIC does well for large samples while AICc tends to be better for small samples where the relative number of parameters is large.

## Example: Los Angeles Pollution Study, 1970s

```{r}
par(mfrow=c(3,1))
tsplot(cmort, main="Cardiovascular Mortality", col=6, type="o", pch=19, ylab="")
tsplot(tempr, main="Temperature", col=4, type="o", pch=19, ylab="")
tsplot(part, main="Particulates", col=2, type="o", pch=19, ylab="")
```

## Example: Los Angeles Pollution Study, 1970s

```{r}
tsplot(cmort, ylab="", ylim=c(20,130), col=astsa.col(6,.8))
lines(tempr, col=astsa.col(4,.9))
lines(part, col=astsa.col(2,.8))
legend("topright", legend=c("Mortality", "Temperature", "Pollution"), lty=1, lwd=2, col=c(6,4,2), bg="white")
```

## Example: Los Angeles Pollution Study, 1970s

- These time series are from a study on the possible effects of temperature and pollution on weekly mortality in Los Angeles County during the 1970s

- We notice strong seasonal components in all 3 time series

- There's a slight downward trend in the cardiovascular mortality in the 10 year period

- There appears to be an inverse relationship between mortality and temperature (mortality rate is higher for cooler temperatures; lower for warmer temps)

- There also appears to be a possible direct relationship between particulates (pollution) and mortality

## Example: Los Angeles Pollution Study, 1970s

To investigate further, we look at the correlation between each of the variables in pairs, and plot their scatterplots also.

```{r}
panel.cor <- function(x, y, ...){
par(usr = c(0, 1, 0, 1))
r <- round(cor(x, y), 2)
text(0.5, 0.5, r, cex = 1.75)
}
pairs(cbind(Mortality=cmort, Temperature=tempr, Particulates=part), col=4, lower.panel=panel.cor)
```

## Example: Los Angeles Pollution Study, 1970s

Observations from this plot...

>

- We see the positive correlation between Mortality and Particulates (pollution); and this correlation looks pretty linear in the scatterplot

>

- We see the negative correlation between Mortality and Temperature; but this correlation does NOT look linear (it curves down and then slightly up again)

>

- Temperature and Particulates are nearly uncorrelated

## Example: Los Angeles Pollution Study, 1970s

Introducing some notation for the 3 different time series:

Let $M_t$ = cardiovascular mortality

Let $T_t$ = temperature

Let $P_t$ = particulate levels

>

We want to try to model $M_t$ using a linear regression. From our analysis so far it seems clear that both $T_t$ and $P_t$ are influencing $M_t$. But we're going to look at several models for comparison.

## Example: Los Angeles Pollution Study, 1970s

Let's think about that non-linearity we saw in the correlation between Mortality and Temperature...

- Because of this we'd like to include a quadratic factor ($T_t^2$) instead of just a linear one ($T_t$).

- But given the temperature range we're looking at, $T_t$ and $T_t^2$ are highly collinear

```{r, echo=TRUE}
par(mfrow=2:1)
plot(tempr, tempr^2)
cor(tempr, tempr^2)
```

## Example: Los Angeles Pollution Study, 1970s

To fix this, we'll adjust for the mean temperature. 

Let $T_{.}$ denote the mean temperature (which for this data is 74.26)

We'll look at variables ($T_t - T_{.}$) and $(T_t - T_{.})^2$ instead. Note they are NOT collinear in this range:

```{r, echo=TRUE}
par(mfrow=2:1)
temp = tempr-mean(tempr)
plot(temp, temp^2)
cor(temp,temp^2)
```

## Example: Los Angeles Pollution Study, 1970s

So, given this, we'll test the following 4 models:

\begin{align}
M_t &= \beta_0 + \beta_1 t + w_t \\
M_t &= \beta_0 + \beta_1 t + \beta_2 (T_t - T_{.}) + w_t \\
M_t &= \beta_0 + \beta_1 t + \beta_2 (T_t - T_{.}) + \beta_3 (T_t - T_{.})^2 + w_t \\
M_t &= \beta_0 + \beta_1 t + \beta_2 (T_t - T_{.}) + \beta_3 (T_t - T_{.})^2 + \beta_4 P_t + w_t 
\end{align}

$(1)$ is a trend only model

$(2)$ adds a linear temperature term

$(3)$ includes the curvilinear (quadratic) temperature term

$(4)$ adds the pollution term

## Example: Los Angeles Pollution Study, 1970s

Here some R code to define the different variables...and fit the different models

```{r, echo=TRUE}
t = time(cmort);          #  time (pure trend)
Tadj = tempr-mean(tempr); #  (T_t - T_.)  temp adjusted by mean
Tadj2 = Tadj^2;           #  (T_t - T_.)^2 
Pt = part;                #  Particulates (pollution)
#
#
fit1 <- lm(cmort ~ t, na.action=NULL)
fit2 <- lm(cmort ~ t + Tadj, na.action=NULL)
fit3 <- lm(cmort ~ t + Tadj + Tadj2, na.action=NULL)
fit4 <- lm(cmort ~ t + Tadj + Tadj2 + Pt, na.action=NULL)


summary(fit1)
summary(fit2)
summary(fit3)
summary(fit4)

anova(fit1)
mean(tempr)
```


## Example: Los Angeles Pollution Study, 1970s

\begin{block}{Exercise 3b.2a}
Try some of the {\tt lm()} commands and then look at the results using the {\tt summary(fit)} and {\tt anova(fit)}.
\vskip .6cm
Note that in the {\tt anova()} output, SSE is the "Sum Sq" of the Residuals and MSE is the "Mean Sq" of the Residuals.
\end{block}

## Example: Los Angeles Pollution Study, 1970s

R has built-in functions to compute AIC and BIC from the output of an **lm()** command but they differ a bit from the definition of AIC and BIC our book uses. Here's some code that extracts the data and makes the needed adjustment:

```{r, echo=TRUE}
num = length(cmort);
t = time(cmort); 
fit1 <- lm(cmort ~ t, na.action=NULL)
AIC(fit3)/num - log(2*pi)
BIC(fit3)/num - log(2*pi)
```

## Example: Los Angeles Pollution Study, 1970s

\begin{block}{Exercise 3b.2b}
Using this methodology, compute the AIC and BIC values for the subsequent models we fit. 
\end{block}

>

Here's a table containing some of the output you should be seeing in your results:

\begin{tabular}{cccccccc} 
Model & $k$ & SSE & df & MSE & $R^2$ & AIC & BIC \\ \hline
$(1)$ & 2 & 40,020 & 506 & 79.0 & .21 & 5.38 & 5.40 \\
$(2)$ & 3 & 31,413 & 505 & 62.2 & .38 & 5.14 & 5.17 \\
$(3)$ & 4 & 27,985 & 504 & 55.5 & .45 & 5.03 & 5.07 \\
$(4)$ & 5 & 20,508 & 503 & 40.8 & .60 & 4.72 & 4.77 
\end{tabular}

## Example: Los Angeles Pollution Study, 1970s

- Examining the table of results we see that model $(4)$ is the "best" since it has the highest $R^2$ with the lowest AIC/BIC values. 

>

- Looking at the output for model $(4)$ we see our "best fit" mortality regression model was:

$$\hat{M}_t = 2831.5 - 1.396t - .472(T_t - 74.26) + 0.023(T_t - 74.26)^2 + .255P_t$$

- Note that:

  - a negative trend is present over time

  - coefficient of adjusted temperature is also negative
  
  - pollution is weighted positively (can be interpreted as incremental contribution to daily deaths per unit of particulate pollution)
  
## Example: Los Angeles Pollution Study, 1970s



- An important final note is that we should really be checking the residuals $\hat{w}_t = M_t - \hat{M}_t$, too...

>

- It turns out there is a fair amount of autocorrelation in these particular residuals (i.e., they are not "white"), which is an issue we'll return to later in the course.

>

- Try: **plot(resid(fit4), type="o")** to see a plot of the residuals for the final model.

## Example: Regression with Lagged Variables

Recall our correlation analysis of the SOI (Southern Oscillation Index) vs Recruitment (new fish) which indicated that the SOI at time $t-6$ months was associated/correlated with Recruitment at time $t$. 

We'll dig deeper into this later but for now let's try fitting a naive linear regression model:

$$R_t = \beta_0 + \beta_1 S_{t-6} + w_t.$$

The R library **dynlm** fits dynamic linear models and time series regressions. For lagged regression, the set-up is faster and more intuitive. 


```{r}
par(mfrow = c(1,2))
plot(resid(fit4), type = "o" )
plot(resid(fit3), type = "o" )
```

> 

Perform the commands below to fit the lagged regression for SOI and Recruitment.

```{r, echo=TRUE, results='hide'}
library(dynlm)
summary(fit <- dynlm(rec ~ L(soi,6)))
```

## Example: Regression with Lagged Variables

```{r}
library(dynlm)
summary(fit <- dynlm(rec ~ L(soi,6)))
```

## Example: Regression with Lagged Variables

The resulting fit model is:

$$\hat{R}_t = 65.79 - 44.28S_{t-6}$$

but taking a look at the residuals, there's a bit too much pattern to think this model is "right"... (stay tuned!)

```{r}
par(mfrow=2:1)
library(dynlm)
fit <- dynlm(rec ~ L(soi,6))
tsplot(resid(fit),col=4)
```
