---
title: "STAT 335 - Problem Set 1"
author: "Ducheng Tan"
date: "2026-01-22"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



\textbf{Problem 1} R contains a dataset called sunspots which contains the monthly mean
relative sunspot numbers from 1749 to 1983, as collected at the Swiss
Federal Observatory, in Zurich (until 1960), and then at the Tokyo
Astronomical Observatory. The astsa library contains a dataset called
sunspotz which contains biannual smoothed (12-month moving average)
number of sunspots from June 1749 to December 1978. Produce two separate
graphs in a multi-figure plot of the sunspots and sunspotz time series.
Make sure your graphs are labelled well enough so that it’s clear which
data is represented in each. Write a short description of what you
observe about these two time series. Include at least three different
observations/features.

```{r}
library(astsa)
```


```{r}
(NROW(sunspots))
(NROW(sunspotz))
par(mfrow = c(2,1))
tsplot(sunspots, col=  "firebrick1", main="sunspots")
tsplot(sunspotz, col= "blue", main = "sunspotz") 
```

The time series plot from the dataset sunspots and sunspotz is plotted above. We can noticed that two plots shows indentical trends, this is due to the fact that the sunpotz is a biannual smoothed version of the sunspots plot. To be exact if $S_t$ is the timeseries for the sunspots data, and $Z_t$ for the sunspotz data, then 

$$Z_t = \frac{1}{12} \sum_{t-6 \leq k \leq t+6} S_k \quad , \quad \forall t \in [1,459]$$
the domain of $t$ is define such that each $S_t$ denotes a unique month in the data. This is true as we have NROW(sunspotz) = 459. 
We can easily see that the sunspotz time series is a smoother version of the sunspots for the reasons we explained above.Overall both timeseries illustrates the same underlying long term patterns in their respective time. 

A: 
1. The sunspots data is much more "jaggy" than the sunspotz data, which makes sense because sunspotz
is a 12-month moving average of the sunspots dataset.
2. There is a clear cyclical/oscillatory pattern in both data sets.
3. There was a dip in the number of sunspots in roughly the 1800-1850s.
4. The largest number of sunspots was around the 1960s.


\textbf{Problem 2}
1.2. There are a number of seismic recordings from earthquakes and from mining explosions in astsa. All of the data are in the dataframe eqexp, but two specific recordings are in EQ5 and EXP6, the fifth earthquake and the sixth explosion, respec-
tively. The data represent two phases or arrivals along the surface, denoted by $P(t=1, ... ,1024)$ and $S (t=1025, ... ,2048)$, at a seismic recording station. The recording instruments are in Scandinavia and monitor a Russian nuclear testing site.
The general problem of interest is in distinguishing between these waveforms in order to maintain a comprehensive nuclear test ban treaty.
To compare the earthquake and explosion signals,
(a) Plot the two series separately in a multifigure plot with two rows and one column.
(b) Plot the two series on the same graph using different colors or different line types.
(c) In what way are the earthquake and explosion series different?

```{r}
#View(eqexp)
#(a)
par(mfrow = c(2, 1))

tsplot(eqexp$EQ5,col="red",main="5th Earthquake")
tsplot(eqexp$EX6,col="blue",main="6th Explosion")

```
```{r}
#(b)
tsplot(EQ5, type = "l", main="EQ5 vs EXP6", col = "red", xlab= "Time Index", ylab = "Amplitude") 
lines(EXP6, type= "l", col="blue", lty = 5)
legend("topright",
legend = c("Earthquake (EQ5)", "Explosion (EXP6)"), col = c("red", "blue"),lty = c(1, 5), lwd = c(2,2))

```
```{r}
ts.plot(ts(eqexp$EQ5), ts(eqexp$EX6), col=c("red","blue"))
legend("topright", legend=c("EQ5","EX6"), lty=1, col=c("red","blue"))
```

(c).
\textbf{Problem 3}
1.3. In this problem, we explore the difference between random walk and moving
average models.

(a) Generate and (multifigure) plot nine series that are random walks (see Exam-
ple 1.10) of length n = 500 without drift ($\delta$ = 0) and $\sigma_w$ = 1.


```{r}
par(mfrow = c(3,3))
for (i in 1:9){
  set.seed(i+3)
  w = rnorm(500); 
  x = cumsum(w)
  t = paste("random walk ", i) 
  tsplot(x, main = t, col = "orange", ylab = "")
}
```


(b) Generate and (multifigure) plot nine series of length n = 500 that are moving
averages of the form (1.1) discussed in Example 1.8.


```{r}
par(mfrow = c(3,3)) 
for (i in 1:9){ 
  set.seed(2*i)
  x = rnorm(500); 
  v = filter(x, sides = 2, filter = rep(1/3, 3));
  t = paste("moving average", i )
  tsplot(v, main = t, ylim=c(-3, 3), col="red", ylab= "")
  
  
  }
```

(c) Comment on the differences between the results of part (a) and part (b).

The moving averages retain the same "shape" as the white noise sequence. The mean is clearly 0 and the
variance is rather tightly controlled. The random walks take off in different directions, some climbing in an
upward trend, others going down into the negative values. There’s a lot more variability in the random walk
series than the moving average ones.


```{r}
w = rnorm(100)
t = 1:100 
x = 3*cos(2*pi*t / 4) + w
v = filter(x, sides = 1 , filter=rep(1/4, 4))
tsplot(x, main="Signal w/Noise and 4-point Moving Average", ylab="Amplitude", col="blue", lty=2);
lines(v, col = "red")
legend("topright", lty = c(2,1), col = c("blue", "red"), bg= "pink", legend = c("x_t", "v_t"))
```
(4.c) What do you observe about the two time series xt and vt in your plot? What are the major features
you see in each? How do the two series compare?
A few observations:
• The Signal w/Noise series has a much larger amplitude than the moving average does.
• The oscillation of the cosine ("signal") is clearly visible, even with the noise (because the amplitude of
the signal is larger than the variance of the noise).
• Both series seem to have mean 0.
• The 4-point moving average seems to smooth out the cosine series a bit too much.
