---
title: "Problem Set 2"
author: "Ducheng Tan"
date: "2026-02-1"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Problem 5 
```{r}
#5a. 
library(astsa)
library(xts)
library(stats)

plot(gtemp_land, col= "blue")


```
I believed that gtemp_land data is non stationary because the expected mean for any time is not the around the same. Also the $cov(x_t, x_s)$ is not constant. 

```{r}
tsplot(cbind(gtemp_land,gtemp_ocean), spaghetti=TRUE, col = astsa.col(c(2,4), .5),
lwd=2, type="o", pch=20, ylab="Temperature Deviations")
legend("topleft", col=c(2,4), lty=1, lwd=2, pch=20, bg="white",
legend=c("Land Surface", "Sea Surface"))

```

Remark : 
Definition 2.13. A stationary time series is a finite variance process where
(i) the mean value function, $\mu_t$ defined in (2.1) is constant and does not depend on

time t, and
(ii) the autocovariance function, $\gamma(s, t)$, defined in (2.2) depends on times $s$ and $t$
only through their time difference.


5b 

```{r}



plot(djia$Close)

```
This again is clearly not stationary as there is no constant  mean and the autocvariance does not appear to only depend on time difference. (Lag)


5c

```{r}
 
x <-rnorm(1000, 5, 6)
plot(x, type = "l" , col = "gold")

```
This is stationary as the mean stays constant around 0, the autocovariance is only dependent on the difference in time. 
Problem 6 

6a. 


$$x_t = \mu + w_t + \theta w_{t-1}, \quad  w_t \sim wn(0, \sigma_w ^2 ) $$
We have $\mathbf{E}(x_t) = \mathbf{E}( w_t + \theta w_{t-1}) = \mu $

6b. Done by hand
6c. 6d. 6e.   

Problem 7 


7a.

```{r}
set.seed(1114)
w <- rnorm(500) 
v = filter(w, side = 2, rep(1/3, 3))
plot(v, col = "orange")

acf = acf(v, lag.max = 20, plot = TRUE, na.action = na.pass) 
acf


```

```{r}

j = rnorm(50)

k = filter(j, rep(1/3, 3), sides = 2)
plot(k, col = "red")

acf_k = acf(k, na.action = na.pass)
acf_k 


```






```{r}
 w = rnorm(50)
v = filter(w, filter=rep(1/3,3), sides=2)
(acf(v, 20, na.action=na.pass))
```




