---
title: "STAT 335, Introduction to Time Series, Unit 1"
subtitle: "(Part c): Time Series Models"
author: "Dr. Mary Lynn Reed"
date: ""
theme: "Warsaw"
colortheme: "wolverine"
output: beamer_presentation
header-includes:
  - \usefonttheme{serif}
  - \newtheorem{prop}{Proposition}
  - \newtheorem{thm}{Theorem}
  - \newtheorem{cor}{Corollary}
  - \newtheorem{exer}{Exercise}
  - \setbeamertemplate{footline}[frame number]
  - \usepackage{setspace}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(xts)
library(astsa)
```

## Modeling Time Series

-   The primary objective of time series analysis is to develop
    mathematical models that provide plausible descriptions for various
    sample time series data



-   From our initial examination of time series data, it's reasonable to
    conclude that many time series data are *correlated* (across time)



-   i.e., the value $x_t$ (at time $t$), may depend in some way on the
    past values $x_{t-1}, x_{t-2}, \ldots$



-   This gives us a start at how to generate realistic looking time
    series data

## White Noise

-   Let $w_t$ denote a series of *uncorrelated* random variables with
    mean 0 and a finite (fixed) variance $\sigma_w^2$



-   In engineering applications, this is often called *white noise*



**Notation:** $w_t \sim wn(0,\sigma_w^2)$



```{=tex}
\vskip .3cm
\begin{block}{}
Why {\bf white}? the term {\it white noise} is used as an analogy to white light, which contains all frequencies in the color spectrum at the same level of intensity.
\end{block}
\vskip .3cm
```


-   One particular kind of white noise is when the variables are
    independent and identically distributed normals; i.e.,
    $w_t \sim \text{iid } N(0,\sigma_w^2)$

## White Noise

```{r, echo=TRUE}
w = rnorm(400) # 400 N(0,1) variates
tsplot(w, col="blue")
```

## White Noise

> -   Notice that this uncorrelated random sample of $N(0,1)$ variates has
    some resemblance to some portions of the DJIA returns




> -   If the stochastic behavior of all time series could be explained in
    terms of the white noise model, classical statistical techniques
    would suffice for time series analysis




> -   But that is not the case, so we need other methods!

## Smoothing

-   The raw white noise samples were quite "jaggy"



> -   Sometimes we want a "smoother" looking series



> -   If we have several $wn(0,\sigma_w^2)$ random variables, what happens when you look at their average?



> -   What statistical result are we leveraging?

## Statistical Reminders

```{=tex}
\begin{block}{Distribution of Sample Mean}
Let $X_1, X_2, \ldots, X_n$ be a random sample from a {\it normal distribution} with mean $\mu$ and standard deviation $\sigma$.
\vskip .4cm
Then for any $n$, the sample mean, $\bar{X}$, is normally distributed with mean $\mu$ and standard deviation $\sigma/\sqrt{n}$.
\vskip .8cm
\end{block}
```


So, in the case of $N(0,1)$, with a sample of size $n$, the sample mean
will be normally distributed with mean 0 and standard deviation
$1/\sqrt{n}$.

## And while we're at it...

-   What if the white noise isn't normal but does have a fixed mean and
    standard deviation; then what can we say about the distribution of its sample means?

```{=tex}
\vskip 1cm
\only<2->{
\begin{block}{Central Limit Theorem}
Let $X_1, X_2, \ldots, X_n$ be a random sample from {\it any distribution} with mean $\mu$ and variance $\sigma^2$. 
\vskip .4cm
Then if $n$ is sufficiently large, the sample mean, $\bar{X}$, has approximately a normal distribution with mean $\mu$ and variance $\sigma^2/n$.
\end{block}
}
```
## Moving Averages

-   Okay, now we have the proper framework to look at a time series
    smoothing technique called \`\`moving averages"

-   Let's replace the white noise series $w_t$ by a \`\`moving average"
    of its values that will produce a smoother time series

-   Specifically, consider
    $$v_t = \frac{1}{3}\left(w_{t-1}+w_t+w_{t+1}\right)$$

## Moving Averages

```{r}
par(mfrow=2:1)
w = rnorm(400) # 400 N(0,1) variates
v = filter(w, sides=2, filter=rep(1/3,3)) # moving average
tsplot(w, col=4, main="white noise")
tsplot(v, ylim=c(-3,3), col=4, main="moving average")
```

## Moving Averages

> -   The moving average sequence is not only smoother but also has
>     lower variance (which we should expect, right?)



> -   What else? (And does it look like any of the data sets we've
>     examined?)



> -   Notice there's some similarity between the moving average series
>     and parts of the fMRI data



> -   What if we want to produce even \`\`smoother" data?

## Trying it with $v_t = \frac{1}{5}\left(w_{t-2}+w_{t-1}+w_t+w_{t+1}+w_{t+2}\right)$

```{r}
par(mfrow=2:1)
w = rnorm(400) # 400 N(0,1) variates
v = filter(w, sides=2, filter=rep(1/5,5)) # moving average
tsplot(w, col=4, main="white noise")
tsplot(v, ylim=c(-3,3), col=4, main="moving average")
```

## In R, this is done with the filter() command

```{=tex}
{\tt v = filter(w, sides=2, filter=rep(1/5,5))}

\vskip .7cm

\begin{block}{Exercise 1c.1a}
Use R help to work out and explain exactly what this statement is doing...
\end{block}

\vskip .7cm

Note what the ``{\tt sides}" parameter does...
```



## Linear Filters

Mathematically, we say that a *linear filter* is a transformation of an
input series $x_t$ using a sequence of coefficients $a_j$ for
$j = 0, \pm 1, \pm 2, \ldots$ as:
$$y_t = \sum_{j=-\infty}^\infty a_j x_{t-j},$$ where
$\sum_{j=-\infty}^\infty |a_j| < \infty.$

This is also sometimes called a *convolution*.


## Moving Average as a Linear Filter

-   Think about moving averages in this context...

>

-   If $v_t = \frac{1}{3}(w_{t-1}+ w_t + w_{t+1})$, then $v_t$ is a
    *linear filter* (or *convolution*) of $w_t$ with what sequence of
    coefficients?
    
>

\only<2->{
{\bf Answer:}
$a_{-1} = a_0 = a_1 = \frac{1}{3}$ and all other $a_j = 0$.
}

## Autoregression

-   Moving averages can create a \`\`smoother" version of a time series
    but it keeps the same basic shape of the data

>

> -   We saw a few different data sets with oscillatory (quasi-periodic)
>     behavior so we need a model that can produce that kind of effect,
>     too

>

> -   Let $w_t$ be a white noise time series.



> -   Consider the following: $$x_t = 1.5x_{t-1} - 0.75x_{t-2} + w_t,$$
>     where we let $x_{-1} = x_0 = 0$.

## Autoregression

-   Given the recursive formula and the initial conditions, the behavior
    at the beginning of the sequence won't have the full oscillatory
    effect

>

-   The \`\`fix" for this is to run the sequence out a while and then
    look at it...

>

Here's some R code that produces the autoregressive sequence (plotted on
next slide)

>

```{r, echo=TRUE}
set.seed(90210)
w = rnorm(250 + 50) # 50 extra to avoid startup problems
x = filter(w, filter=c(1.5,-.75), 
           method="recursive")[-(1:50)]
```

## Autoregression

```{r}
set.seed(90210)
w = rnorm(250 + 50) # 50 extra to avoid startup problems
x = filter(w, filter=c(1.5,-.75), 
           method="recursive")[-(1:50)]
tsplot(x, main="autoregression", col=4)
```

## Autoregression with filter() command

```{=tex}
\begin{block}{Exercise 1c.1b}
To illustrate the workings of the "recursive" filter, look at the following example. Do the following in R, and then look at the values in the variables {\tt x} and {\tt y}:

{\tt x <- 1:10}

{\tt y <- filter(x, filter=c(1,1), method="recursive")}

\vskip .4cm
Now, do the first few calculations of the recurrence:
$$y_t = y_{t-1} + y_{t-2} + x_t$$
by hand, assuming $y_{-1} = y_0 = 0$ and where $x_1 = 1, x_2, = 2, x_3 = 3,$ etc. 
\vskip .5cm
Your $y_t$ sequence should be the values that appear in the {\tt y} variable... 
\end{block}
```
## Random Walk with Drift

> -   So far we've seen how to produce random behavior (white noise),
>     smoothing effects (moving averages), and oscillations
>     (autoregression)

> -   Next we need a model for trend (e.g., the upward climb of a stock
>     price, or Earth's temperature, for example)

> -   E.g., the *random walk with drift* model:
>     $$x_t = \delta + x_{t-1} + w_t$$ where $w_t$ is white noise and
>     $x_0 = 0$ is the initial condition. The constant $\delta$ is
>     called the *drift*.

> -   When $\delta = 0$, this is just called a *random walk*.

## Random Walk with Drift

```{=tex}
\begin{block}{Exercise 1c.2}
Show that the random walk with drift model is equivalent to: 
$$x_t = \delta t + \sum_{j=1}^t w_j$$ 
for $t = 1,2,\ldots$.
\end{block}
```
## Random Walk with Drift

```{r}
set.seed(314159265) # so you can reproduce the results
w  = rnorm(200) 
x  = cumsum(w)
wd = w +.3 
xd = cumsum(wd)
tsplot(xd, ylim=c(-2,80), main="random walk", ylab="", col=4)
 clip(0, 200, 0, 80)
 abline(a=0, b=.3, lty=2, col=4) # drift
lines(x, col=6)
 clip(0, 200, 0, 80)
 abline(h=0, col=6, lty=2)
 legend(0, 80, lty=1, lwd=2, col=c(4,6), bg="white",
        legend=c("rw with drift 0.3", "rw without drift"))
```

## Signal Plus Noise

-   Many models for generating time series assume an underlying
    \`\`signal" with some consistent periodic variation contaminated by
    noise

> -   E.g., consider
>     $$x_t = 2 \cos\left(2\pi \frac{t+15}{50}\right) + w_t,$$ where the
>     first term is considered the *signal* and $w_t$ is white noise

## Signal Plus Noise

```{=tex}
\begin{block}{}
The sinusoidal waveform {\it signal} 
$$2 \cos\left(2\pi \frac{t+15}{50}\right)$$
can be written as: 
$$A\cos(2\pi\omega t + \phi),$$
where $A$ is the amplitude, $\omega$ is the frequency of the oscillation, and $\phi$ is the phase shift. 
\vskip .4cm
What are the constants $A$, $\omega$, and $\phi$ in this example?
\end{block}
```

\pause

Answers: $A = 2$, $\omega = 1/50$, $\phi = 0.6\pi$


## Signal Plus Noise

```{r}
cs = 2*cos(2*pi*(1:500+15)/50)           # same thing 
w  = rnorm(500,0,1)
par(mfrow=c(3,1))   
tsplot(cs, ylab="", col=4, main = expression(x[t]==2*cos(2*pi*t/50+.6*pi)))
tsplot(cs + w, ylab="", col=4, main = expression(x[t]==2*cos(2*pi*t/50+.6*pi)+N(0,1)))
tsplot(cs + 5*w, ylab="", col=4, main = expression(x[t]==2*cos(2*pi*t/50+.6*pi)+N(0,25)))
```

## Signal Plus Noise

-   Using a *signal plus noise* model, the degree to which the signal is
    obscured depends on the amplitude of the signal relative to the size
    of $\sigma_w$.

>

> -   The ratio of the amplitude of the signal to $\sigma_w$ (or some
>     function of the ratio) is sometimes called the *signal-to-noise
>     ratio (SNR)*

>

> -   The larger the SNR, the easier it is to detect the signal

>

> -   In real time series data, we will likely only see the signal
>     obscured by noise

## Putting Two Models Together

```{=tex}
\begin{block}{Exercise 1c.3}
{\bf 1c.3a} Use a favorite number to set a specific seed for the R random
number generator and then generate a white noise sequence
$w_t \sim N(0,1)$ for $t = 1,2,\ldots, 200$.

\vskip .3cm

{\bf 1c.3b} Generate $n = 100$ observations from the autoregression model
$$x_t = 0.55x_{t-2} + w_t,$$ using the methods we considered in class to
avoid start-up problems (but skip the first 100 values instead of just
50).

\vskip .3cm

{\bf 1c.3c} Apply the moving average filter
$$v_t = (x_t + x_{t-1} + x_{t-2} + x_{t-3})/4,$$ to the $x_t$ data you
generated. (Notice this a "past-values only" filter.)
\end{block}
```

## Putting Two Models Together (continued)

```{=tex}
\begin{block}{Exercise 1c.3}
{\bf 1c.3d} On the same graph, plot $x_t$ as a dashed line of one color
and $v_t$ as a solid line of a different color. Include appropriate
labels on the axes, a descriptive legend, and a title for your plot.


\vskip .5cm

{\bf 1c.3e} What do you observe about the two time series $x_t$ and $v_t$
in your plot? What are the major features you see in each? How do the
two series compare?
\end{block}
```
