---
title: "Statistical Machine Learning - Fundamentals - Classification Learning - 1"
author: "Ernest Fokoué"
date: "2024-12-06"
output:
  pdf_document: default
  pdf: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chunk

```{R, echo=FALSE}

  library(MASS)
  library(datasets)
  library(kernlab)
  library(mlbench)
  library(bestglm) 
  library(glmnet)
  library(ggplot2)
  library(BMS)
  library(corrplot)
  library(car)
  library(ROCR)
  
```


## Introduction to the problem and the data

We are first studying the famous Pima Indian Diabetes data originating from the area of Phoenix in Arizona. This data featuring 768 females of Pima Indian ancestry for which it become important to not only classify but also comprehend the risk factors for diabetes!

```{R, echo=FALSE}
  
   data("PimaIndiansDiabetes")
   #View(PimaIndiansDiabetes)


# Load the PimaIndiansDiabetes dataset
data(PimaIndiansDiabetes)

# Store the dataset in a data frame
df <- PimaIndiansDiabetes

# First 6 rows
head(df)

# Last 6 rows
tail(df)

# Sample size
n <- nrow(df)
cat("Sample size (n):", n, "\n")

# Dimensionality of input space (excluding the response variable)
p <- ncol(df) - 1
cat("Dimensionality of input space (p):", p, "\n")

# Position of the response variable
py <- which(names(df) == "diabetes")
cat("Position of the response variable (py):", py, "\n")

# Check for missing values
missing_count <- sum(is.na(df))
cat("Number of missing values:", missing_count, "\n")

# Declare richness if n/p > 5
richness <- ifelse(n / p > 5, "richness", "not richness")
cat("Richness status:", richness, "\n")

# Rename response variable to "Y"
XY <- df
names(XY)[names(XY) == "diabetes"] <- "Y"

# Display first and last 6 rows of the modified dataset
head(XY)
tail(XY)

```

## Exploration of the Response Variable 

Let's now explore the distribution of the response variable along with the idiosyncratic assessment of the individual discriminative and predictive power of each predictor (explanatory) variable.

```{R, echo=FALSE}

# Frequency and percentage distributions
response_freq <- table(df$diabetes)
response_perc <- prop.table(response_freq) * 100

# Print the distributions
cat("Frequency distribution:\n")
print(response_freq)
cat("\nPercentage distribution:\n")
print(response_perc)

# Create beautiful barplots
par(mfrow = c(1, 2))
barplot(response_freq, main = "Frequency Distribution", col = "skyblue", xlab = "Response", ylab = "Frequency")
barplot(response_perc, main = "Percentage Distribution", col = "lightgreen", xlab = "Response", ylab = "Percentage")

# Compute max(freq)/min(freq)
imbalance_ratio <- max(response_freq) / min(response_freq)
imbalance_status <- ifelse(imbalance_ratio > 10, "severe imbalance", "reasonable balance")

cat("Max/Min Ratio:", imbalance_ratio, "\n")
cat("Imbalance Status:", imbalance_status, "\n")

# Compute Kruskal-Wallis test statistics for each predictor
kruskal_results <- sapply(names(df)[-which(names(df) == "diabetes")], function(var) {
  kruskal.test(as.formula(paste(var, "~ diabetes")), data = df)$statistic
})

# Store results as a named vector
names(kruskal_results) <- names(df)[-which(names(df) == "diabetes")]

# Plot the results using type "h"
par(mfrow = c(1, 1))
plot(kruskal_results, type = "h", main = "Kruskal-Wallis Test Statistics", xlab = "Predictor Variables",
     ylab = "Test Statistic", col = "blue", xaxt = "n")
axis(1, at = 1:length(kruskal_results), labels = names(kruskal_results), las = 2)

# Identify best and worst predictors
best_predictor <- names(kruskal_results)[which.max(kruskal_results)]
worst_predictor <- names(kruskal_results)[which.min(kruskal_results)]

cat("Best Predictor:", best_predictor, "\n")
cat("Worst Predictor:", worst_predictor, "\n")

# Set up multi-panel plotting
par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))

# Create comparative boxplots
for (var in names(df)[-which(names(df) == "diabetes")]) {
  boxplot(as.formula(paste(var, "~ diabetes")), data = df, main = paste("Boxplot of", var),
          col = c("pink", "lightblue"), xlab = "Response", ylab = var)
}

# Set up multi-panel plotting
par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))

# Create comparative histograms
for (var in names(df)[-which(names(df) == "diabetes")]) {
  hist(df[[var]][df$diabetes == "pos"], main = paste("Histogram of", var, "(pos)"),
       xlab = var, col = "red", breaks = 10, freq = TRUE)
  hist(df[[var]][df$diabetes == "neg"], main = paste("Histogram of", var, "(neg)"),
       xlab = var, col = "blue", breaks = 10, freq = TRUE, add = TRUE)
  legend("topright", legend = c("pos", "neg"), fill = c("red", "blue"))
}

```

## Correlation analysis and input in-depth look

```{R, echo=FALSE}

   str(XY[,-py]) # What are the types of the input variables

   # Do we need any transformation like standardizing the input

   boxplot(XY[,-py], col=1:p)
   
   corrplot(cor(XY[,-py]))
```

## Introduction to Fitting logistic Regression

```{R, echo=FALSE}

  logistic.xy <- glm(Y~., data=XY, family=binomial(link='logit'))
  print(summary(logistic.xy))
```

## Chunk

```{R, echo=FALSE}

# Build the logistic regression model
model <- glm(diabetes ~ ., data = PimaIndiansDiabetes, family = binomial(link = 'logit'))

# Print model summary
summary(model)

# Compute the Chi-Squared test
chi_sq <- model$null.deviance - model$deviance
df <- model$df.null - model$df.residual
overall_pvalue <- pchisq(chi_sq, df, lower.tail = FALSE)

cat("Chi-Squared Statistic:", chi_sq, "\n")
cat("Degrees of Freedom:", df, "\n")
cat("Overall P-value:", overall_pvalue, "\n")

# Compute Pseudo R²
pseudo_r2 <- 1 - (model$deviance / model$null.deviance)
cat("Pseudo R²:", pseudo_r2, "\n")


```

## Chunk

```{R, echo=FALSE}
# Perform stepwise selection using AIC
aic_model <- stats::step(model, scope=~., trace=0, direction = "both", k=2)

# Print summary
summary(aic_model)

# Plot residuals
par(mfrow = c(1, 2))
plot(residuals(aic_model), main = "Residuals of AIC Model", col = "blue")
hist(residuals(aic_model), main = "Histogram of Residuals", col = "lightblue")

```

## Chunk

```{R, echo=FALSE}
# Perform stepwise selection using BIC
bic_model <- stats::step(model, direction = "both", k = log(nrow(PimaIndiansDiabetes)))

# Print summary
summary(bic_model)

# Plot residuals
par(mfrow = c(1, 2))
plot(residuals(bic_model), main = "Residuals of BIC Model", col = "green")
hist(residuals(bic_model), main = "Histogram of Residuals", col = "lightgreen")

```

## Chunk

```{R, echo=FALSE}
# Load necessary library
library(bestglm)

# Prepare data for bestglm (convert response to 0/1)
data_bestglm <- PimaIndiansDiabetes
data_bestglm$diabetes <- ifelse(data_bestglm$diabetes == "pos", 1, 0)

# Perform best subset selection
best_subset <- bestglm(data_bestglm, IC = "BIC", family = binomial)

# Print best subset model
print(best_subset)

```

## Chunk

```{R, echo=FALSE}
# Load necessary libraries
library(glmnet)

# Prepare data for glmnet
X <- model.matrix(diabetes ~ . - 1, data = PimaIndiansDiabetes)
y <- ifelse(PimaIndiansDiabetes$diabetes == "pos", 1, 0)

# Perform cross-validated LASSO
cv_lasso <- cv.glmnet(X, y, family = "binomial", alpha = 1)

# Fit LASSO model at the best lambda
lasso_model <- glmnet(X, y, family = "binomial", alpha = 1, lambda = cv_lasso$lambda.min)

# Plot the solution path
plot(cv_lasso$glmnet.fit, xvar = "lambda", label = TRUE, sign.lambda=+1)
title("LASSO Solution Path")

```


## Chunk

```{R, echo=FALSE}
# Load necessary libraries
library(glmnet)

# Prepare data for glmnet
X <- model.matrix(diabetes ~ . - 1, data = PimaIndiansDiabetes)
y <- ifelse(PimaIndiansDiabetes$diabetes == "pos", 1, 0)

# Perform cross-validation to find the optimal lambda
cv_lasso <- cv.glmnet(X, y, family = "binomial", alpha = 1)

# Extract the best lambda
best_lambda <- cv_lasso$lambda.min
cat("Optimal Lambda from Cross-Validation:", best_lambda, "\n")

# Fit LASSO model using the best lambda
lasso_model <- glmnet(X, y, family = "binomial", alpha = 1, lambda = best_lambda)

# Print the coefficients of the LASSO model
cat("LASSO Coefficients at Optimal Lambda:\n")
print(coef(lasso_model))

# Plot the LASSO solution path
plot(cv_lasso$glmnet.fit, xvar = "lambda", sign.lambda=+1, label = TRUE, main = "LASSO Solution Path")

# Overlay labels for non-zero coefficients
coefficients <- coef(cv_lasso$glmnet.fit, s = best_lambda)
non_zero <- which(coefficients != 0)  # Indices of non-zero coefficients
variable_names <- rownames(coefficients)[non_zero]

if (length(non_zero) > 0) {
  for (i in seq_along(non_zero)) {
    index <- non_zero[i]
    # Log-transform lambda values and label only non-zero coefficients
    text(log(cv_lasso$lambda), 
         coefficients[index, ], 
         labels = variable_names[i], cex = 0.7, pos = 4, col = "red")
  }
}

```


## Chunk

```{R, echo=FALSE}

# Load necessary libraries
library(MASS)

# Prepare data for logistic regression
data <- PimaIndiansDiabetes

# Fit the full logistic regression model
full_model <- glm(diabetes ~ ., data = data, family = binomial)

# Perform stepwise selection using BIC
bic_model <- stepAIC(full_model, direction = "both", trace=0, k = log(nrow(data)))

# Print the summary of the BIC model
summary(bic_model)

# Extract the coefficients from the BIC model
coefficients <- coef(bic_model)

# Compute the relative change in odds for each predictor
# Relative change in odds = (exp(coefficient) - 1) * 100
odds_change <- (exp(coefficients) - 1) * 100

# Print the relative changes
cat("Relative Change in Odds (in %):\n")
print(odds_change)

# Interpretation for glucose
cat("For glucose: A one-unit increase leads to a", round(odds_change["glucose"], 2), 
    "% change in the odds of diabetes.\n")

# Interpretation for pregnant
cat("For pregnant: A one-unit increase leads to a", round(odds_change["pregnant"], 2), 
    "% change in the odds of diabetes.\n")

```



## Logistic Regression beyond the interpretation: Prediction and Performance assessment

```{R, echo=FALSE}

# Fit the full logistic regression model
full_model <- glm(diabetes ~ ., data = PimaIndiansDiabetes, family = binomial)

# Perform stepwise selection using BIC
bic_model <- stepAIC(full_model, direction = "both", trace=0, k = log(nrow(PimaIndiansDiabetes)))

# Extract and print the variables in the BIC model
cat("Variables included in the BIC model:\n")
print(names(coef(bic_model)))

```


## Chunk

```{R, echo=FALSE}
# Fit the final logistic regression model using BIC-selected variables
logistic.xy.final <- glm(diabetes ~ pregnant + glucose + mass + pedigree, data = PimaIndiansDiabetes, 
                         family = binomial)

# Print summary of the model
summary(logistic.xy.final)

# (a) Predict probabilities
predicted_probs <- predict(logistic.xy.final, type = "response")

# (b) Predict labels (threshold = 0.5)
predicted_labels <- ifelse(predicted_probs > 0.5, "pos", "neg")

# (c) Confusion Matrix
true_labels <- PimaIndiansDiabetes$diabetes
conf_matrix <- table(True = true_labels, Predicted = predicted_labels)
cat("Confusion Matrix:\n")
print(conf_matrix)


# (d) Plot predicted probabilities
plot(predicted_probs, 
     col = ifelse(true_labels == "pos", "red", "blue"), 
     pch = 16, xlab = "Index", ylab = "Predicted Probability", 
     main = "Predicted Probabilities")
legend("topright", legend = c("Positive", "Negative"), 
       col = c("red", "blue"), pch = 16)

library(ROCR)

# (e) Plot ROC curve
pred <- prediction(predicted_probs, true_labels)
perf <- performance(pred, "tpr", "fpr")
plot(perf, col = "blue", lwd = 2, main = "ROC Curve")
abline(a = 0, b = 1, lty = 2, col = "gray") # Diagonal line

# (f) Compute AUC
auc <- performance(pred, "auc")@y.values[[1]]
cat("Area Under the Curve (AUC):", round(auc, 4), "\n")

# (g) Compute evaluation metrics
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
error_rate <- 1 - accuracy
tpr <- conf_matrix["pos", "pos"] / sum(conf_matrix["pos", ])
fpr <- conf_matrix["neg", "pos"] / sum(conf_matrix["neg", ])
precision <- conf_matrix["pos", "pos"] / sum(conf_matrix[, "pos"])
recall <- tpr
f_score <- 2 * (precision * recall) / (precision + recall)

# Create a table for evaluation metrics
metrics <- data.frame(
  Metric = c("Accuracy", "Error Rate", "True Positive Rate (TPR)", 
             "False Positive Rate (FPR)", "Precision", "Recall", "F-Score"),
  Value = c(accuracy, error_rate, tpr, fpr, precision, recall, f_score)
)
print(metrics)

# Declare results
cat("Model Evaluation Metrics:\n")
print(metrics)


```

## Predictive Analytics Full Scale

# Chunk

```{R, echo=FALSE}
# Load required libraries
library(MASS)         # For stepAIC (AIC and BIC models)
library(class)        # For kNN
library(rpart)        # For classification tree
library(ROCR)         # For ROC curves
library(dplyr)        # For data manipulation

# Load the PimaIndiansDiabetes dataset
data("PimaIndiansDiabetes", package = "mlbench")

# Define helper functions for metrics
compute_metrics <- function(conf_matrix) {
  accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
  error <- 1 - accuracy
  precision <- conf_matrix["pos", "pos"] / sum(conf_matrix[, "pos"])
  recall <- conf_matrix["pos", "pos"] / sum(conf_matrix["pos", ])
  tpr <- recall
  fpr <- conf_matrix["neg", "pos"] / sum(conf_matrix["neg", ])
  f_score <- 2 * precision * recall / (precision + recall)
  
  return(c(accuracy, error, precision, recall, tpr, fpr, f_score))
}
```



# Subchunk

```{R, echo=FALSE}
set.seed(123)  # For reproducibility
S <- 100       # Number of splits
train_ratio <- 0.7  # Training set proportion

# Storage for metrics
metrics_list <- list(train = list(), test = list())

# Loop over stochastic splits
for (i in 1:S) 
{
  # Split data
  train_index <- sample(1:nrow(PimaIndiansDiabetes), 
                        size = train_ratio * nrow(PimaIndiansDiabetes))
  train_set   <- PimaIndiansDiabetes[train_index, ]
  test_set    <- PimaIndiansDiabetes[-train_index, ]
  
  # True labels
  true_train <- train_set$diabetes
  true_test  <- test_set$diabetes
  
  # (a) AIC logistic regression
  aic_model <- stepAIC(glm(diabetes ~ ., data = train_set, family = binomial), 
                       direction = "both", trace = FALSE)
  pred_train_aic <- ifelse(predict(aic_model, type = "response") > 0.5, "pos", "neg")
  pred_test_aic <- ifelse(predict(aic_model, newdata = test_set, type = "response") > 0.5, "pos", "neg")
  
  # (b) BIC logistic regression
  bic_model <- stepAIC(glm(diabetes ~ ., data = train_set, family = binomial), 
                       direction = "both", k = log(nrow(train_set)), trace = FALSE)
  pred_train_bic <- ifelse(predict(bic_model, type = "response") > 0.5, "pos", "neg")
  pred_test_bic <- ifelse(predict(bic_model, newdata = test_set, type = "response") > 0.5, "pos", "neg")
  
  # (c) 1-Nearest Neighbor
  pred_train_knn <- knn(train_set[, -9], train_set[, -9], cl = true_train, k = 1)
  pred_test_knn <- knn(train_set[, -9], test_set[, -9], cl = true_train, k = 1)
  
  # (d) Classification Tree
  tree_model <- rpart(diabetes ~ ., data = train_set, method = "class")
  pred_train_tree <- predict(tree_model, type = "class")
  pred_test_tree <- predict(tree_model, newdata = test_set, type = "class")
  
  # Compute metrics
  conf_train_aic <- table(True = true_train, Predicted = pred_train_aic)
  conf_test_aic <- table(True = true_test, Predicted = pred_test_aic)
  
  conf_train_bic <- table(True = true_train, Predicted = pred_train_bic)
  conf_test_bic <- table(True = true_test, Predicted = pred_test_bic)
  
  conf_train_knn <- table(True = true_train, Predicted = pred_train_knn)
  conf_test_knn <- table(True = true_test, Predicted = pred_test_knn)
  
  conf_train_tree <- table(True = true_train, Predicted = pred_train_tree)
  conf_test_tree <- table(True = true_test, Predicted = pred_test_tree)
  
  metrics_list$train[[i]] <- list(
    AIC = compute_metrics(conf_train_aic),
    BIC = compute_metrics(conf_train_bic),
    kNN = compute_metrics(conf_train_knn),
    Tree = compute_metrics(conf_train_tree)
  )
  
  metrics_list$test[[i]] <- list(
    AIC = compute_metrics(conf_test_aic),
    BIC = compute_metrics(conf_test_bic),
    kNN = compute_metrics(conf_test_knn),
    Tree = compute_metrics(conf_test_tree)
  )
}

```


# SubChunk

```{R, echo=FALSE}
# ROC curves for the last split
last_pred_train_aic <- prediction(predict(aic_model, type = "response"), true_train)
last_pred_test_aic <- prediction(predict(aic_model, newdata = test_set, type = "response"), true_test)

last_pred_train_bic <- prediction(predict(bic_model, type = "response"), true_train)
last_pred_test_bic <- prediction(predict(bic_model, newdata = test_set, type = "response"), true_test)

last_pred_train_knn <- prediction(as.numeric(pred_train_knn) - 1, as.numeric(true_train) - 1)
last_pred_test_knn <- prediction(as.numeric(pred_test_knn) - 1, as.numeric(true_test) - 1)

last_pred_train_tree <- prediction(as.numeric(predict(tree_model, type = "prob")[, 2]), as.numeric(true_train) - 1)
last_pred_test_tree <- prediction(as.numeric(predict(tree_model, newdata = test_set, type = "prob")[, 2]), as.numeric(true_test) - 1)

# Performance objects
perf_train_aic <- performance(last_pred_train_aic, "tpr", "fpr")
perf_test_aic <- performance(last_pred_test_aic, "tpr", "fpr")

perf_train_bic <- performance(last_pred_train_bic, "tpr", "fpr")
perf_test_bic <- performance(last_pred_test_bic, "tpr", "fpr")

perf_train_knn <- performance(last_pred_train_knn, "tpr", "fpr")
perf_test_knn <- performance(last_pred_test_knn, "tpr", "fpr")

perf_train_tree <- performance(last_pred_train_tree, "tpr", "fpr")
perf_test_tree <- performance(last_pred_test_tree, "tpr", "fpr")

# Plotting ROC curves
plot(perf_train_aic, col = "blue", lwd = 2, main = "ROC Curves (Last Split)")
lines(perf_test_aic@x.values[[1]], perf_test_aic@y.values[[1]], col = "blue", lty = 2)

lines(perf_train_bic@x.values[[1]], perf_train_bic@y.values[[1]], col = "green", lwd = 2)
lines(perf_test_bic@x.values[[1]], perf_test_bic@y.values[[1]], col = "green", lty = 2)

lines(perf_train_knn@x.values[[1]], perf_train_knn@y.values[[1]], col = "red", lwd = 2)
lines(perf_test_knn@x.values[[1]], perf_test_knn@y.values[[1]], col = "red", lty = 2)

lines(perf_train_tree@x.values[[1]], perf_train_tree@y.values[[1]], col = "purple", lwd = 2)
lines(perf_test_tree@x.values[[1]], perf_test_tree@y.values[[1]], col = "purple", lty = 2)

# Adding legend
legend("bottomright", inset = 0.02, 
       legend = c("AIC Train", "AIC Test", "BIC Train", "BIC Test", 
                  "kNN Train", "kNN Test", "Tree Train", "Tree Test"),
       col = c("blue", "blue", "green", "green", "red", "red", "purple", "purple"),
       lty = c(1, 2, 1, 2, 1, 2, 1, 2), lwd = 2)


```


## Chunk

```{R, echo=FALSE}
# Set up plotting space for two panels
par(mfrow = c(1, 2))  # One row, two columns

# Training ROC curves
plot(perf_train_aic, col = "blue", lwd = 2, main = "Training ROC Curves")
lines(perf_train_bic@x.values[[1]], perf_train_bic@y.values[[1]], col = "green", lwd = 2)
lines(perf_train_knn@x.values[[1]], perf_train_knn@y.values[[1]], col = "red", lwd = 2)
lines(perf_train_tree@x.values[[1]], perf_train_tree@y.values[[1]], col = "purple", lwd = 2)

# Add legend for Training
legend("bottomright", inset = 0.02, 
       legend = c("AIC", "BIC", "kNN", "Tree"),
       col = c("blue", "green", "red", "purple"),
       lty = 1, lwd = 2)

# Test ROC curves
plot(perf_test_aic, col = "blue", lwd = 2, main = "Test ROC Curves")
lines(perf_test_bic@x.values[[1]], perf_test_bic@y.values[[1]], col = "green", lwd = 2)
lines(perf_test_knn@x.values[[1]], perf_test_knn@y.values[[1]], col = "red", lwd = 2)
lines(perf_test_tree@x.values[[1]], perf_test_tree@y.values[[1]], col = "purple", lwd = 2)

# Add legend for Test
legend("bottomright", inset = 0.02, 
       legend = c("AIC", "BIC", "kNN", "Tree"),
       col = c("blue", "green", "red", "purple"),
       lty = 1, lwd = 2)

```


## Get the errors

```{r cross_validation_loop, echo=FALSE}
# Monte Carlo Cross-Validation Loop with S iterations
S <- 100  # Number of Monte Carlo iterations
test_errors <- matrix(0, nrow = S, ncol = 4)  # Store test errors for 4 models
colnames(test_errors) <- c("AIC", "BIC", "kNN", "Tree")

set.seed(123)  # For reproducibility

for (s in 1:S) 
{
  # Split data into training and test sets
  train_indices <- sample(1:nrow(attitude), size = 0.7 * nrow(attitude))
  train_data    <- attitude[train_indices, ]
  test_data     <- attitude[-train_indices, ]
  
  # 1. AIC model
  aic_model <- stats::step(lm(rating ~ ., data = train_data), 
                    scope = ~., trace = 0, direction = "both", k = 2)
  aic_pred  <- predict.lm(aic_model, newdata = test_data)
  test_errors[s, "AIC"] <- mean((test_data$rating - aic_pred)^2)
  
  # 2. BIC model  
  bic_model <- stats::step(lm(rating ~ ., data = train_data),
                    scope = ~., trace = 0, direction = "both", k = log(nrow(train_data)))
  bic_pred <- predict.lm(bic_model, newdata = test_data)
  test_errors[s, "BIC"] <- mean((test_data$rating - bic_pred)^2)
  
  # 3. kNN model
  library(class)
  # Standardize features for kNN
  train_std <- scale(train_data[, -1])
  test_std  <- scale(test_data[, -1], 
                    center = attr(train_std, "scaled:center"),
                    scale = attr(train_std, "scaled:scale"))
  
  k <- 5  # Number of neighbors
  knn_pred <- knn(train_std, test_std, train_data$rating, k = k)
  test_errors[s, "kNN"] <- mean((test_data$rating - as.numeric(as.character(knn_pred)))^2)
  
  # 4. Decision Tree
  library(rpart)
  tree_model <- rpart(rating ~ ., data = train_data, method = "anova")
  tree_pred  <- predict(tree_model, newdata = test_data)
  test_errors[s, "Tree"] <- mean((test_data$rating - tree_pred)^2)
  
  # Progress indicator
  if (s %% 20 == 0) cat("Completed", s, "iterations\n")
}

# Display summary statistics
cat("\nAverage Test Errors:\n")
print(colMeans(test_errors))
cat("\nStandard Deviations:\n")
print(apply(test_errors, 2, sd))

# Now you can use test_errors for plotting
# For example: boxplot(test_errors, main = "Test Errors by Model")
```

## Prepare and boxplot the errors

```{R, echo=FALSE}
# Convert test errors to a data frame for plotting
test_error_df <- as.data.frame(t(test_errors))
colnames(test_error_df) <- c("AIC", "BIC", "kNN", "Tree")

# Plot boxplots for test errors
boxplot(test_error_df, 
        col = rainbow(4), 
        main = "Test Errors by Model", 
        xlab = "Model", 
        ylab = "Test Error")

```

## Getting some F-scores

```{r fscore_cross_validation, echo=FALSE}
# Monte Carlo Cross-Validation Loop with S iterations for F-scores
# For classification tasks (e.g., using Pima Indians Diabetes dataset)

# Load Pima Indians Diabetes dataset
library(mlbench)
data(PimaIndiansDiabetes)

# Use diabetes dataset for classification
data <- PimaIndiansDiabetes
S <- 100  # Number of Monte Carlo iterations
test_f_scores <- matrix(0, nrow = S, ncol = 4)  # Store F-scores for 4 models
colnames(test_f_scores) <- c("Logistic_AIC", "Logistic_BIC", "kNN", "Tree")

# Function to calculate F1 score
calculate_f1 <- function(predicted, actual, positive_class = "pos") {
  # Confusion matrix components
  actual_pos <- actual == positive_class
  predicted_pos <- predicted == positive_class
  
  tp <- sum(actual_pos & predicted_pos)
  fp <- sum(!actual_pos & predicted_pos)
  fn <- sum(actual_pos & !predicted_pos)
  
  # Calculate precision, recall, and F1
  precision <- ifelse((tp + fp) > 0, tp / (tp + fp), 0)
  recall <- ifelse((tp + fn) > 0, tp / (tp + fn), 0)
  f1 <- ifelse((precision + recall) > 0, 
               2 * precision * recall / (precision + recall), 
               0)
  
  return(f1)
}

set.seed(123)  # For reproducibility

for (s in 1:S) {
  # Split data into training and test sets
  train_indices <- sample(1:nrow(data), size = 0.7 * nrow(data))
  train_data    <- data[train_indices, ]
  test_data     <- data[-train_indices, ]
  
  # 1. Logistic Regression with AIC selection
  full_model <- glm(diabetes ~ ., data = train_data, family = binomial)
  aic_model <- stats::step(full_model, scope = ~., trace = 0, direction = "both", k = 2)
  aic_pred_prob <- predict(aic_model, newdata = test_data, type = "response")
  aic_pred <- ifelse(aic_pred_prob > 0.5, "pos", "neg")
  test_f_scores[s, "Logistic_AIC"] <- calculate_f1(aic_pred, test_data$diabetes)
  
  # 2. Logistic Regression with BIC selection
  bic_model <- stats::step(full_model, scope = ~., trace = 0, direction = "both", 
                    k = log(nrow(train_data)))
  bic_pred_prob <- predict(bic_model, newdata = test_data, type = "response")
  bic_pred <- ifelse(bic_pred_prob > 0.5, "pos", "neg")
  test_f_scores[s, "Logistic_BIC"] <- calculate_f1(bic_pred, test_data$diabetes)
  
  # 3. kNN model
  library(class)
  # Standardize numeric features for kNN
  numeric_cols   <- sapply(train_data, is.numeric)
  train_features <- as.matrix(train_data[, numeric_cols])
  test_features  <- as.matrix(test_data[, numeric_cols])
  
  train_std <- scale(train_features)
  test_std  <- scale(test_features, 
                    center = attr(train_std, "scaled:center"),
                    scale = attr(train_std, "scaled:scale"))
  
  k <- 5  # Number of neighbors
  knn_pred <- knn(train_std, test_std, train_data$diabetes, k = k)
  test_f_scores[s, "kNN"] <- calculate_f1(knn_pred, test_data$diabetes)
  
  # 4. Decision Tree
  library(rpart)
  tree_model <- rpart(diabetes ~ ., data = train_data, method = "class")
  tree_pred  <- predict(tree_model, newdata = test_data, type = "class")
  test_f_scores[s, "Tree"] <- calculate_f1(tree_pred, test_data$diabetes)
  
  # Progress indicator
  if (s %% 20 == 0) cat("Completed", s, "F-score iterations\n")
}

# Display summary statistics
cat("\nAverage F1 Scores (higher is better):\n")
print(colMeans(test_f_scores))
cat("\nStandard Deviations:\n")
print(apply(test_f_scores, 2, sd))

# Create boxplot
boxplot(test_f_scores, 
        main = "F1 Scores by Model",
        ylab = "F1 Score",
        col = c("lightblue", "lightgreen", "lightcoral", "lightgoldenrod"),
        ylim = c(0, 1))
grid()

```

## Showing the F-scores

```{R, echo=FALSE}
# Convert test F-scores to a data frame for plotting
test_f_score_df <- as.data.frame(t(test_f_scores))
colnames(test_f_score_df) <- c("AIC", "BIC", "kNN", "Tree")

# Plot boxplots for F-scores
boxplot(test_f_score_df, 
        col = rainbow(4), 
        main = "Test F-scores by Model", 
        xlab = "Model", 
        ylab = "F-Score")

```

# Subchunk

```{R, echo=FALSE}
library(class)
library(rpart)
library(mlbench)

set.seed(123)

# Initialize metrics list
metrics_list <- list(
  AIC = list(train = vector("list", S), test = vector("list", S)),
  BIC = list(train = vector("list", S), test = vector("list", S)),
  kNN = list(train = vector("list", S), test = vector("list", S)),
  CART = list(train = vector("list", S), test = vector("list", S))
)

# Helper function to compute performance metrics
compute_metrics <- function(predicted, actual) {
  actual <- factor(actual)
  predicted <- factor(predicted, levels = levels(actual))
  cm <- table(Predicted = predicted, Actual = actual)
  accuracy <- sum(diag(cm)) / sum(cm)
  error <- 1 - accuracy
  precision <- diag(cm) / rowSums(cm)
  recall <- diag(cm) / colSums(cm)
  f_score <- 2 * (precision * recall) / (precision + recall)
  fpr <- colSums(cm - diag(cm)) / sum(cm)
  tpr <- recall
  
  metrics <- c(Accuracy = accuracy, Error = error, Precision = mean(precision, na.rm = TRUE),
               Recall = mean(recall, na.rm = TRUE), TPR = mean(tpr, na.rm = TRUE), 
               FPR = mean(fpr, na.rm = TRUE), F_score = mean(f_score, na.rm = TRUE))
  return(metrics)
}

# Main loop for S stochastic holdouts
for (s in 1:S) {
  # Split the data into training and test sets
  train_indices <- sample(1:nrow(PimaIndiansDiabetes), 
                          size = 0.7 * nrow(PimaIndiansDiabetes))
  test_indices  <- setdiff(1:nrow(PimaIndiansDiabetes), train_indices)
  
  train_data    <- PimaIndiansDiabetes[train_indices, ]
  test_data     <- PimaIndiansDiabetes[test_indices, ]
  
  # AIC logistic regression
  aic_model <- stats::step(glm(diabetes ~ ., data = train_data, 
                               family = binomial(link = "logit")), 
                           direction = "both", trace = FALSE)
  aic_train_preds <- predict(aic_model, train_data, type = "response")
  aic_test_preds  <- predict(aic_model, test_data, type = "response")
  metrics_list$AIC$train[[s]] <- compute_metrics(ifelse(aic_train_preds > 0.5, "pos", "neg"), train_data$diabetes)
  metrics_list$AIC$test[[s]] <- compute_metrics(ifelse(aic_test_preds > 0.5, "pos", "neg"), test_data$diabetes)
  
  # BIC logistic regression
  bic_model <- stats::step(glm(diabetes ~ ., data = train_data, 
                        family = binomial(link = "logit")), 
                        direction = "both", k = log(nrow(train_data)), 
                        trace = FALSE)
  bic_train_preds <- predict(bic_model, train_data, type = "response")
  bic_test_preds  <- predict(bic_model, test_data, type = "response")
  metrics_list$BIC$train[[s]] <- compute_metrics(ifelse(bic_train_preds > 0.5, "pos", "neg"), train_data$diabetes)
  metrics_list$BIC$test[[s]] <- compute_metrics(ifelse(bic_test_preds > 0.5, "pos", "neg"), test_data$diabetes)
  
  # kNN classification
  knn_train_preds <- knn(train = train_data[, -9], 
                         test = train_data[, -9], cl = train_data$diabetes, k = 1)
  knn_test_preds  <- knn(train = train_data[, -9], 
                        test = test_data[, -9], cl = train_data$diabetes, k = 1)
  metrics_list$kNN$train[[s]] <- compute_metrics(knn_train_preds, train_data$diabetes)
  metrics_list$kNN$test[[s]] <- compute_metrics(knn_test_preds, test_data$diabetes)
  
  # CART classification
  cart_model <- rpart(diabetes ~ ., data = train_data, method = "class")
  cart_train_preds <- predict(cart_model, train_data, type = "class")
  cart_test_preds <- predict(cart_model, test_data, type = "class")
  metrics_list$CART$train[[s]] <- compute_metrics(cart_train_preds, train_data$diabetes)
  metrics_list$CART$test[[s]] <- compute_metrics(cart_test_preds, test_data$diabetes)
}

# Aggregate metrics
average_metrics_train <- sapply(metrics_list, function(x) colMeans(do.call(rbind, x$train), na.rm = TRUE))
average_metrics_test <- sapply(metrics_list, function(x) colMeans(do.call(rbind, x$test), na.rm = TRUE))

# Plot average training and test errors
plot(1:4, average_metrics_train["Error", ], type = "b", col = "blue", ylim = c(0, max(average_metrics_test["Error", ], average_metrics_train["Error", ], na.rm = TRUE)),
     xlab = "Machine", ylab = "Error", main = "Average Training and Test Errors", xaxt = "n")
lines(1:4, average_metrics_test["Error", ], type = "b", col = "red")
axis(1, at = 1:4, labels = c("AIC", "BIC", "kNN", "CART"))
legend("bottomleft", inset=0.02, legend = c("Training Error", "Test Error"), col = c("blue", "red"), lty = 1)


```

# Subchunk

```{R, echo=FALSE}
# Prepare the data for the table

AverageTraining <- average_metrics_train[2,]
AverageTest     <- average_metrics_test[2,]

average_errors_table <- data.frame(
  Model = c("Logistic Regression (AIC)", "Logistic Regression (BIC)", 
            "k-Nearest Neighbors", "Classification Tree"),
  Avg_Training_Error = AverageTraining,
  Avg_Test_Error = AverageTest
)

# Load knitr for table rendering
library(knitr)

# Display the table
kable(average_errors_table, 
      caption = "Average Training and Test Errors for Each Model",
      col.names = c("Model", "Avg Training Error", "Avg Test Error"),
      digits = 4,  # Adjust precision for better readability
      align = "lcc")  # Align columns: left for Model, center for errors

```


We are always Happy and Joyful! Thank you!
