---
title: "HW4-Computations"
author: "Ducheng Tan"
date: "2026-02-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
---
title: "STAT 406 Computations"
author: "Ducheng Tan"
date: "2026-01-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}


a_book = (qf(0.95, 10,5))
1/a_book

a_me = qf(0.05,5, 10)
a_me

b_book = qf(0.95, 5, 10)
b_book

p_a= pf(a_me, 5, 10)
p_b = pf(b_book, 5, 10)
p_b - p_a

par(mfrow = c(1,2))

x = seq(0, 10, 0.001); plot(df(x,10,5)~x, ylab="F(r_1 = 10, r_2 = 5)", xlab = "x", main = "F(r_1 = 10, r_2 = 5)", col= "orange")

y = seq(0, 10, 0.001); plot(df(1/y ,5, 10)*(1/y^2) ~ y, ylab="X ~ F(r_1 = 5, r_2 = 10)", xlab = "x", main = "1/X ~ F(r_1 = 5, r_2 = 10)", col= "red")

```

```{r}
y = seq(0.01, 10, 0.01)
plot(df(y, 10, 5) ~ y, type="l", 
     ylab="Density", xlab="F", 
     main="F ~ F(5, 10)", 
     col="blue", lwd=2)

# 1/F distribution: transform with Jacobian
# If Y = 1/F, then f_Y(y) = f_F(1/y) * |dF/dY| = f_F(1/y) * (1/y^2)
y = seq(0.01, 10, 0.01)
density_1_over_F = df(1/y, 5, 10) * (1/y^2)
plot(density_1_over_F ~ y, type="l", 
     ylab="Density", xlab="1/F", 
     main="1/F where F ~ F(5, 10)", 
     col="red", lwd=2)
```


```{r}
library(pwr)

find_sample_size <- function(means, sigma2, alpha = 0.05, target_power = 0.90) {
  
  a <- length(means)
  
  
  mu_bar <- mean(means)
  
 
  tau <- means - mu_bar
  

  sum_tau2 <- sum(tau^2)
  
  max_diff <- max(means) - min(means)
  
  cat("=== Input Summary ===\n")
  cat("Number of groups (a):          ", a, "\n")
  cat("Group means:                   ", means, "\n")
  cat("Overall mean:                  ", mu_bar, "\n")
  cat("Treatment effects (tau_i):     ", tau, "\n")
  cat("Sum of squared effects:        ", sum_tau2, "\n")
  cat("Maximum difference:            ", max_diff, "\n")
  cat("Error variance (sigma^2):      ", sigma2, "\n")
  cat("Alpha:                         ", alpha, "\n")
  cat("Target power:                  ", target_power, "\n")
  cat("\n=== Searching for n ===\n")
  

for (n in 2:200) {
    
    
    phi2 <- (n * sum_tau2) / (a * sigma2)
    phi  <- sqrt(phi2)
    
  
    df1 <- a - 1          
    df2 <- a * (n - 1)    
    N   <- a * n          
    

    F_crit <- qf(1 - alpha, df1, df2)
    
    ncp   <- (n * sum_tau2) / sigma2
    power <- pf(F_crit, df1, df2, ncp = ncp, lower.tail = FALSE)
    
    cat(sprintf("n = %3d | Phi^2 = %6.3f | Phi = %5.3f | df2 = %3d | Power = %.4f\n",
                n, phi2, phi, df2, power))
    
    if (power >= target_power) {
      cat("\n=== Result ===\n")
      cat("Required sample size per group: n =", n, "\n")
      cat("Total observations needed:      N =", N, "\n")
      cat("Achieved power:                    ", round(power, 4), "\n")
      cat("Phi^2:                             ", round(phi2, 4), "\n")
      cat("Phi:                               ", round(phi,  4), "\n")
      return(invisible(list(n = n, N = N, power = power, 
                            phi2 = phi2, phi = phi,
                            df1 = df1, df2 = df2)))
    }
  }
  cat("No Solution for this power level")
}
means         <- c(50, 60, 50, 60)
sigma2        <- 25
alpha         <- 0.05
target_power  <- 0.90

result <- find_sample_size(means, sigma2, alpha, target_power)
```



```

